Incorporating periodic variability in hidden Markov models for animal movement

### 2016-07-05 paper3.Rnw
temp push

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
\todo{it occurs to me that we haven't really included any
information about simulation testing in this version!
We should probably add a short section; if don't, then we
should delete following sentence}
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

\todo{maybe give the formula for number of parameters (constrained) for each model type here?}

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data (\todo{fill in value} $> ?? $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
fit of each model type.
(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
captures serial correlation at both short and long scales. 
We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
L <- load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

Figure~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. 
\todo{is it \textbf{exactly} consistent? Weibull/vM/homog gets 5, not 6? Can we explain?}
Furthermore, the BIC-optimal number of states does not depend on the model within the time-homogenous or time-heterogeneous frameworks, but time-heterogeneous models have the BIC-optimal number of states overall.\todo{?? clarify? do you mean time-het models have lower overall BIC (not shown in this graph)?}

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
L <- load("cat1sumdat.RData")
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+
    facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="# of parameters") +
    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin)
@

>>>>

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
\todo{it occurs to me that we haven't really included any
information about simulation testing in this version!
We should probably add a short section; if don't, then we
should delete following sentence }
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

\todo{maybe give the formula for number of parameters (constrained) for each model type here?...ML: Sum(I+T+E), I=Prior=n(s-1) where n is the number of parameters for the prior complexity, T = ns(s-1), E=2s}

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
fit of each model type.
(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
captures serial correlation at both short and long scales. 
We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
#L <- load("cat1allmod.RData")
allmodsumdf2 <- readRDS("cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-heterogeneous HMM","homogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

Figure~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles) 
\todo{is it \textbf{exactly} consistent? Weibull/vM/homog gets 5, not 6? Can we explain?}

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(grepl("HMM",model),"no","yes")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model,linetype=model))+
    facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="# of parameters") +
    scale_x_continuous(expand=c(0,0.5),breaks=2:7)+
    geom_point(aes(size=parameters))+
    scale_linetype_manual(values=linetypev)+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin)
@
